---
title: "Biostatistics, Biology Masters, Meeting 3-1; Independent Work"
author: "Adam Clark"
output: html_document
---

## Homework problem:

Try fitting different structures of models (combinations of fixed effect slopes and intercepts, and random effects vs. nested random effects).

```{r echo=TRUE}
bdat = read.csv("dataset_3.1.csv")
n_plots = length(unique(bdat$plot))

# plot
par(mar=c(4,4,2,2))
plot(biomass~diversity, col = rainbow(n_plots)[plot], pch = subplot, data = bdat)
```

Note - different point types show different subplots, and different colors show different plots. We can plot these trends a bit more informatively using the `coplot` command. It is set up similarly to the `lme` function (with some extra bits that we'll discuss later).

```{r echo=TRUE}
par(mar=c(4,4,2,2))
coplot(biomass~diversity|as.factor(plot), data = bdat,
       panel = function(x, y, ...) {points(x,y, pch = bdat$subplot[bdat$plot==1])})
```

What form of regression should you use? Random slopes? Random intercepts? Nested? Something else?

What do the results of the regression show us? What are the fixed and random effects estimates? What is the variability across parameters?